Bayesian Hierarchical Modeling: An Introduction and Reassessment

### 2023-06-30 manuscript.Rmd
Updates (replaced and removed files)

\Revision{Up until now, we specified models without a correlation between the individual effects. This means, for instance, that we do not expect that the estimations for the individual effects $\delta_{7, i}$ and $\delta_{3, i}$ to be related. However, in cases where we would expect the estimation of individual effects to be related, we could estimate the correlation coefficient between individual effects. This would add complexity to the model. For the sake of the tutorial, we will focus on the models without individual effect correlation, but we illustrate the estimation of the model with correlation between the individual effects in} [Online Supplement C and D](https://github.com/MyrtheV/Bayesian-Hierarchical-Modelling-An-Introduction-and-Reassessment).

@@

```{r distritypeplot, echo=FALSE, fig.cap="Common probability density distributions that may be used as priors. Panel A shows normal distributions varying in mean and standard deviation. Panel B shows inverse gamma distributions that differ in the shape and scale parameter. Panel C shows Student's \\textit{t}-distributions that differ in the location and scale parameter. For all three \\textit{t}-distributions, the degrees of freedom are set to 3.", fig.align = "left", fig.height = 8, fig.width = 12, out.width="100%"}
knitr::include_graphics("I - Images/Fig4_distrtypefigure_v3.png")
```

@@

```{r priorpredfigurepanel, echo=FALSE, fig.cap="Side and digit effects from predicted data. Violins represent the distribution of predicted effects across 1000 simulation runs. The blue points are the average effects across simultions. Panel A shows the prior predictions for the normal model based on the default priors of \\textit{brms}. Panel B shows the prior predictions for the normal model based on the selected priors. Panel C shows the prior predictions for the log-normal model based on the priors chosen for the normal model. Panel D shows the prior predictions for the log-normal model based on the selected priors.", fig.align = "left", fig.height = 3, fig.width = 4, out.width="100%"}
knitr::include_graphics("I - Images/Fig5_priorpredfigure_v2.png")
```

@@

```{r traceplots3p, echo=FALSE, fig.align = "left", fig.height = 11, fig.width = 4, out.width="100%", fig.cap="Panel A and B show the trace plot of the digit parameter $\\mu_{\\delta_{6}}$ for both packages. Panel C shows the estimated $\\hat{R}$ for all individual effects per parameter group. The stars represent the $\\hat{R}$ of the general effects. Panel D illustrates the variation in the estimated number of effective samples per parameter group. The dashed lines represent the total number of iterations for each package. The stars represent the number of effective samples of the general effects."}
knitr::include_graphics("I - Images/Fig8_traceplots3pnew_v2.png")
```

@@

```{r modelposteriorsmufigure3, echo=FALSE, fig.cap="Panel A shows the empirical mean RTs per digit. The standard error is represented by bars around the dots. Panel B and C show the posterior distributions for the general effects estimated by the two packages. The middle line within the distributions represents the posterior mean. The shaded area within the distributions represent 95\\% credible intervals.", fig.align = "left", fig.height = 11, fig.width = 1, out.width="100%"}
knitr::include_graphics("I - Images/Fig9_posteriormu_v3.png")
```

@@

```{r randomparfigure, echo=FALSE, fig.cap="The posterior means for individual effect parameters with 95\\% credible intervals as estimated by \\textit{rstan}, shown in increasing order. The dashed line represents the general posterior mean. Pink intervals contain zero, blue interval do not contain zero.", fig.align = "left", fig.height = 1, fig.width = 0.5, out.width="100%"}
knitr::include_graphics("I - Images/Fig10_randompar_v2.png")

@@

```{r modelestimatesfigure3, echo=FALSE, fig.cap="Model estimates for digit effect parameters in \\textit{rstan}. The points represent the mean parameter estimates for each individual. The violin plots on the right side show the variance of the individual parameter estimates. ", fig.align = "left", fig.height = 11, fig.width = 1, out.width="75%"}
knitr::include_graphics("I - Images/Fig11_modelestimatefigure2_onlyrstan_v2.png")
```

@@

```{r traceplotlog, echo=FALSE, fig.cap="(ref:captiontraceplotlog)", fig.align = "left", fig.height = 11, fig.width = 4, out.width="100%"}
knitr::include_graphics("I - Images/Fig12_traceplots3plog_v2.png")
```

### Results
#### General effects  
We inspect the posterior distributions for the general effect parameters presented in Figure \@ref(fig:postplotlog). 
```{r postplotlog, echo=FALSE, fig.cap="The posterior distributions for the general effects in the log-normal model estimated by the two packages. The middle line within the distributions represents the posterior mean. The shaded area within the distributions represent 95\\% credible intervals.", fig.align = "left", fig.height = 11, fig.width = 1, out.width="100%"}
knitr::include_graphics("I - Images/Fig13_posteriormulog_v2.png")
```

@@

```{r randomparplotlog, echo=FALSE, fig.cap="The posterior means for individual effect parameters in the log-normal model with 95\\% credible intervals as estimated by \\textit{rstan}, shown in increasing order. The dashed line represents the general posterior mean. Pink intervals contain zero, blue interval do not contain zero.", fig.align = "left", fig.height = 1, fig.width = 0.5, out.width="100%"}
knitr::include_graphics("I - Images/Fig14_randomparlog_v2.png")
```
We can further inspect the individual estimates for the digit effects in Figure \@ref(fig:modeldigitestplotlog). The figure shows the individual estimates for the digit effects with the distribution of the variability of the individual estimates (not the posterior distribution of the general effect). The figure illustrates that there is considerable variance in the point estimates of the individual digit effects. However, compared to the normal model there is considerably more hierarchical shrinkage. \Revision{This means that in the log-normal model, the individual effects are corrected more towards the group mean, compared to the normal model. Or, the group mean in the log-normal model is not affected as much by individuals showing a divergent trend compared to the normal model. Whether we should prefer a model with more shrinkage depends on the theoretical considerations. If more individual variation is expected, less shrinkage would be preferred. However, the advantage of more shrinkage is that outliers do not influence the estimation of the effect as much. }

@@

```{r modeldigitestplotlog, echo=FALSE, fig.cap="Model estimates for digit effect parameters in the log-normal model in \\textit{rstan}. The points represent the mean parameter estimates for each individual. The violin plots on the right side show the variance of the individual parameter estimates.", fig.align = "left", fig.height = 11, fig.width = 1, out.width="75%"}
knitr::include_graphics("I - Images/Fig15_inddiflogplot_onlyrstan_v2.png")
```

@@

```{r sdratioplotall, echo=FALSE, fig.cap="The prior and posterior distribution of the side parameter estimated by the \\textit{brms} package. The dots represent the height of the distributions at zero. The ratio of these dots indicates the Savage-Dickey density ratio. The Bayes factor represents the evidence for the hypothesis that the effect equals zero. A Bayes factor above one indicates that the evidence is in favor of the null hypothesis.", fig.align = "left", fig.height = 6, fig.width = 4, out.width="60%"}
knitr::include_graphics("I - Images/Fig16_sdrplotside_v2.png")
```
With `rstan`, the SDD-ratio can be computed by fitting a density function to the posterior distribution obtained from the MCMC samples, and comparing the value of this density function at $\mu_{\beta} = 0$ to the value of the prior distribution at $\mu_{\beta} = 0$. This is illustrated in the code block below. First, we extract the posterior samples for the parameter $\mu_{\beta}$, the general effect of side. We apply the `logspline` function [@logspline] to the samples to obtain an estimate of the log-density of the posterior distribution. The density at the zero is calculated with the `dlogspline` function. Finally, we compare the estimated density of the posterior to the density of the prior. \textcolor{blue}{This process is explained in further detail by Van Ravenzwaaij and Etz} [-@van2021simulation].

@@

```{r bsbfestimatesbrmsrstan, echo=FALSE, fig.cap="The BF estimates of the other models (0; null, side and digit model) against the full model (1). The green triangles represent the estimates using \\textit{rstan}. The orange dots represent the \\textit{brms} estimates. The dashed line represents a $\\text{BF}_{01}$ of 1. ", fig.align = "left", fig.height = 11, fig.width = 4, out.width="70%"}
knitr::include_graphics("I - Images/Fig17_normalbfplot01.png")
```
Finally, it is time to answer our research question: Is there a symbolic distance effect? When considering all Bayes factor estimates from Figure \@ref(fig:sdratioplotall) and \@ref(fig:bsbfestimatesbrmsrstan), the null model ($\text{BF}_{max} = 232 * 10^{105}$ for `rstan`) performs best compared to the side model ($\text{BF}_{max} = 458 * 10^{68}$ for `rstan`), the digit model ($\text{BF}_{max} = 269 * 10^{31}$ for `rstan`). Therefore, the data from Rouder et al. (2005) show evidence against the symbolic distance effect.

@@

### \Revision{Log-normal Model}
\Revision{We can take a similar approach for the log-normal model to compare the hierarchical models on varying complexity. We evaluate their performance in two scenarios that correspond to scenario 1 and 3 for the normal model. This means that in scenario 1 for the log-normal model, we estimated a null, digit, side and full model based on the prior specified in Equation, specifying large individual differences. For scenario 2, we remove individual variability from the side, digit, and full model. This resulted in the following posterior model probabilities shown in Table} \@ref(tab:postmodproblog).

@@

The examples illustrate the benefits of comparing hierarchical models of varying complexity. While individual differences are a common phenomenon in psychology, there are instances where general effects are sufficient to describe the data adequately, or where only a random intercept, but no random slopes are necessary. As shown in Tables} \@ref(tab:postmodprob) and \@ref(tab:postmodproblog), Bayesian model comparisons can help researchers to identify the model that best describes the data.

>>>>

\Revision{Up until now, we specified models without a correlation between the individual effects. This means, for instance, that we do not expect the estimations for the individual effects $\delta_{7, i}$ and $\delta_{3, i}$ to be related. However, in cases where we would expect the estimation of individual effects to be related, we could estimate the correlation coefficient between individual effects. This would add complexity to the model. For the sake of the tutorial, we will focus on the models without individual effect correlation, but we illustrate the estimation of the model with correlation between the individual effects in} [Online Supplement C and D](https://github.com/MyrtheV/Bayesian-Hierarchical-Modelling-An-Introduction-and-Reassessment).

@@

```{r distritypeplot, echo=FALSE, fig.cap="Common probability density distributions that may be used as priors. Panel A shows normal distributions varying in mean and standard deviation. Panel B shows inverse gamma distributions that differ in the shape and scale parameter. Panel C shows Student's \\textit{t}-distributions that differ in the location and scale parameter. For all three \\textit{t}-distributions, the degrees of freedom are set to 3.", fig.align = "left", fig.height = 8, fig.width = 12, out.width="100%"}
knitr::include_graphics("I - Images/Fig5_distrtypefigure_v3.png")
```

@@

```{r priorpredfigurepanel, echo=FALSE, fig.cap="Side and digit effects from predicted data. Violins represent the distribution of predicted effects across 1000 simulation runs. The blue points are the average effects across simultions. Panel A shows the prior predictions for the normal model based on the default priors of \\textit{brms}. Panel B shows the prior predictions for the normal model based on the selected priors. Panel C shows the prior predictions for the log-normal model based on the priors chosen for the normal model. Panel D shows the prior predictions for the log-normal model based on the selected priors.", fig.align = "left", fig.height = 3, fig.width = 4, out.width="100%"}
knitr::include_graphics("I - Images/Fig6_priorpredfigure_v2.png")
```

@@

```{r traceplots3p, echo=FALSE, fig.align = "left", fig.height = 11, fig.width = 4, out.width="100%", fig.cap="Panel A and B show the trace plot of the digit parameter $\\mu_{\\delta_{6}}$ for both packages. Panel C shows the estimated $\\hat{R}$ for all individual effects per parameter group. The stars represent the $\\hat{R}$ of the general effects. Panel D illustrates the variation in the estimated number of effective samples per parameter group. The dashed lines represent the total number of iterations for each package. The stars represent the number of effective samples of the general effects."}
knitr::include_graphics("I - Images/Fig9_traceplots3pnew_v2.png")
```

@@

```{r modelposteriorsmufigure3, echo=FALSE, fig.cap="Panel A shows the empirical mean RTs per digit. The standard error is represented by bars around the dots. Panel B and C show the posterior distributions for the general effects estimated by the two packages. The middle line within the distributions represents the posterior mean. The shaded area within the distributions represent 95\\% credible intervals.", fig.align = "left", fig.height = 11, fig.width = 1, out.width="100%"}
knitr::include_graphics("I - Images/Fig10_posteriormu_v3.png")
```

@@

```{r randomparfigure, echo=FALSE, fig.cap="The posterior means for individual effect parameters with 95\\% credible intervals as estimated by \\textit{rstan}, shown in increasing order. The dashed line represents the general posterior mean. Pink intervals contain zero, blue interval do not contain zero.", fig.align = "left", fig.height = 1, fig.width = 0.5, out.width="100%"}
knitr::include_graphics("I - Images/Fig11_randompar_v2.png")

@@

```{r modelestimatesfigure3, echo=FALSE, fig.cap="Model estimates for digit effect parameters in \\textit{rstan}. The points represent the mean parameter estimates for each individual. The violin plots on the right side show the variance of the individual parameter estimates. ", fig.align = "left", fig.height = 11, fig.width = 1, out.width="75%"}
knitr::include_graphics("I - Images/Fig12_modelestimatefigure2_onlyrstan_v2.png")
```

@@

```{r traceplotlog, echo=FALSE, fig.cap="(ref:captiontraceplotlog)", fig.align = "left", fig.height = 11, fig.width = 4, out.width="100%"}
knitr::include_graphics("I - Images/Fig13_traceplots3plog_v2.png")
```

### Results
#### General effects  
We inspect the posterior distributions for the general effect parameters presented in Figure \@ref(fig:postplotlog). 
```{r postplotlog, echo=FALSE, fig.cap="The posterior distributions for the general effects in the log-normal model estimated by the two packages. The middle line within the distributions represents the posterior mean. The shaded area within the distributions represent 95\\% credible intervals.", fig.align = "left", fig.height = 11, fig.width = 1, out.width="100%"}
knitr::include_graphics("I - Images/Fig14_posteriormulog_v2.png")
```

@@

```{r randomparplotlog, echo=FALSE, fig.cap="The posterior means for individual effect parameters in the log-normal model with 95\\% credible intervals as estimated by \\textit{rstan}, shown in increasing order. The dashed line represents the general posterior mean. Pink intervals contain zero, blue interval do not contain zero.", fig.align = "left", fig.height = 1, fig.width = 0.5, out.width="100%"}
knitr::include_graphics("I - Images/Fig15_randomparlog_v2.png")
```
We can further inspect the individual estimates for the digit effects in Figure \@ref(fig:modeldigitestplotlog). The figure shows the individual estimates for the digit effects with the distribution of the variability of the individual estimates (not the posterior distribution of the general effect). The figure illustrates that there is considerable variance in the point estimates of the individual digit effects. However, compared to the normal model there is considerably more hierarchical shrinkage. \Revision{This means that in the log-normal model, the individual effects are corrected more towards the group mean, compared to the normal model. Or, the group mean in the log-normal model is not affected as much by individuals showing a divergent trend compared to the normal model. Whether we should prefer a model with more shrinkage depends on the theoretical considerations. If more individual variation is expected, less shrinkage would be preferred. However, the advantage of more shrinkage is that outliers do not influence the estimation of the effect as much. }

@@

```{r modeldigitestplotlog, echo=FALSE, fig.cap="Model estimates for digit effect parameters in the log-normal model in \\textit{rstan}. The points represent the mean parameter estimates for each individual. The violin plots on the right side show the variance of the individual parameter estimates.", fig.align = "left", fig.height = 11, fig.width = 1, out.width="75%"}
knitr::include_graphics("I - Images/Fig16_inddiflogplot_onlyrstan_v2.png")
```

@@

```{r sdratioplotall, echo=FALSE, fig.cap="The prior and posterior distribution of the side parameter estimated by the \\textit{brms} package. The dots represent the height of the distributions at zero. The ratio of these dots indicates the Savage-Dickey density ratio. The Bayes factor represents the evidence for the hypothesis that the effect equals zero. A Bayes factor above one indicates that the evidence is in favor of the null hypothesis.", fig.align = "left", fig.height = 6, fig.width = 4, out.width="60%"}
knitr::include_graphics("I - Images/Fig17_sdrplotside_v2.png")
```
With `rstan`, the SDD-ratio can be computed by fitting a density function to the posterior distribution obtained from the MCMC samples, and comparing the value of this density function at $\mu_{\beta} = 0$ to the value of the prior distribution at $\mu_{\beta} = 0$. This is illustrated in the code block below. First, we extract the posterior samples for the parameter $\mu_{\beta}$, the general effect of side. We apply the `logspline` function [@logspline] to the samples to obtain an estimate of the log-density of the posterior distribution. The density at the zero is calculated with the `dlogspline` function. Finally, we compare the estimated density of the posterior to the density of the prior. \textcolor{blue}{This process is explained in further detail by Van Ravenzwaaij and Etz} [-@van2021simulation].

@@

```{r bsbfestimatesbrmsrstan, echo=FALSE, fig.cap="The BF estimates of the other models (0; null, side and digit model) against the full model (1). The green triangles represent the estimates using \\textit{rstan}. The orange dots represent the \\textit{brms} estimates. The dashed line represents a $\\text{BF}_{01}$ of 1. ", fig.align = "left", fig.height = 11, fig.width = 4, out.width="70%"}
knitr::include_graphics("I - Images/Fig18_normalbfplot01.png")
```
Finally, it is time to answer our research question: Is there a symbolic distance effect? When considering all Bayes factor estimates from Figure \@ref(fig:sdratioplotall) and \@ref(fig:bsbfestimatesbrmsrstan), the null model ($\text{BF}_{max} = 232 * 10^{105}$ for `rstan`) performs best compared to the side model ($\text{BF}_{max} = 458 * 10^{68}$ for `rstan`), the digit model ($\text{BF}_{max} = 269 * 10^{31}$ for `rstan`). Therefore, the data from Rouder et al. (2005) show evidence against the symbolic distance effect.

@@

### \Revision{Log-normal Model}
\Revision{We can take a similar approach for the log-normal model to compare the hierarchical models on varying complexity. We evaluate their performance in two scenarios that correspond to scenario 1 and 3 for the normal model. This means that in scenario 1 for the log-normal model, we estimated a null, digit, side and full model based on the prior specified in Equation 8, specifying large individual differences. For scenario 2, we remove individual variability from the side, digit, and full model. This resulted in the following posterior model probabilities shown in Table} \@ref(tab:postmodproblog).

@@

The examples illustrate the benefits of comparing hierarchical models of varying complexity. While individual differences are a common phenomenon in psychology, there are instances where general effects are sufficient to describe the data adequately, or where only a random intercept, but no random slopes are necessary. As shown in Tables \@ref(tab:postmodprob) and \@ref(tab:postmodproblog), Bayesian model comparisons can help researchers to identify the model that best describes the data.
